---
title: "Interacting with APIs: Covid19api"
author: "Yan Liu"
date: "2021/9/30"
output: 
   github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is a vignette to show how to retrieve data from an [API](https://covid19api.com/). To demonstrate, I'll be interacting with the API. I'm going to build a few functions to interact with some of the endpoints and explore some of the data I can retrieve.

As a note, some of these functions return data at a team level. Some of the APIs use the franchise ID number, while some use the most recent team ID to select a specific team's endpoint. For that reason, if you use any of my functions I recommend supplying a full team name (e.g. `"Montréal Canadiens"`). My functions will decode them to the appropriate ID number.

# Requirements

To use the functions for interacting with the NHL API, I used the following packages:

- [`tidyverse`](https://www.tidyverse.org/): tons of useful features for data manipulation and visualization
- [`jsonlite`](https://cran.r-project.org/web/packages/jsonlite/): API interaction

In addition to those packages, I used the following packages in the rest of the document:

- [`cowplot`](https://cran.r-project.org/web/packages/cowplot/index.html): extra functionality for `ggplot2`
- [`imager`](https://cran.r-project.org/web/packages/imager/): loading in images
- [`broom`](https://cran.r-project.org/web/packages/broom/vignettes/broom.html):
tidying up a regression output for display
- [`knitr`](https://cran.r-project.org/web/packages/knitr/index.html): displaying tables in a markdown friendly way
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)
library(dbplyr)
library(rmarkdown)
library(lubridate)
```

# API Interaction Functions

Here is where I define the functions to interact with the [Country Names](https://api.covid19api.com/countries) and [Covid cases summary](https://api.covid19api.com/summary) as well as some helper functions.

## `Find country names`
Pattern match of country names. Multiple names can be a input as a vector.
```{r f1, echo=FALSE}
match_country_name <- function (name){
  #get the list of available countries
  countries <- fromJSON("https://api.covid19api.com/countries")
  country_names<-countries$Country
  name_all <- countries[grep(paste(name,collapse="|"), country_names,ignore.case=TRUE),]
  return(name_all)
}
```

## `Total cases`
Total confirmed cases and deaths today from selected countries, sorted by confirmed cases from high to low.
```{r f2, echo=FALSE}
summary_total <- function(country = "all"){
     # Get data from the summary endpoint
    summary <- fromJSON("https://api.covid19api.com/summary")
     # Select the data part from JSON
    all <- summary$Countries
    
    if (country[1] != "all"){
   
    # match input country names to get their unique Slug name
    country_slug <- match_country_name(country)$Slug
  
    if(length(country_slug) == 0){
      message <- paste("ERROR: Argument for country was not found.",
                       "Try summary_total('all') to find the country you're looking for.")
      stop(message)
    }
    else {
      # Filter the selected countries
      select_country <- all %>% 
      filter(Slug %in% country_slug) %>% 
      select(Country,TotalConfirmed,TotalDeaths,Date) %>% 
      mutate(Date = as_date(Date)) %>% 
      arrange(desc(TotalConfirmed))
    }
  }
  
  else {
    select_country <- all %>% 
      select(Country,TotalConfirmed,TotalDeaths,Date) %>% 
      mutate(Date = as_date(Date)) %>% 
      arrange(desc(TotalConfirmed))
  }
  return(select_country)
}
```

## `New cases`
New cases and death by days in a given period of one country, enter start and end date as YYYY-MM-DD.
```{r f3, echo=FALSE}
new_period <- function(country,start,end){
  
  # match input country names to get their unique Slug name
  country_slug <- match_country_name(country)$Slug
  
  
  # Error message for invalid input of country names
  if(length(country_slug) == 0){
    message <- paste("ERROR: Argument for country was not found. Try summary_total('all') to find the country you're looking for:",
                     )
    stop(message)
  }
  
  # if country name has multiple hits, ask the user select one from the list of possible county names
  else if(length(country_slug) >1){
    message <- paste("ERROR: Argument for country was not unique.",
                     "Please select one from the list:",
                     paste(grep(country, country_names,ignore.case=TRUE,value=TRUE),collapse = '; '))
    stop(message)
  }
  
  else {
    # Get data from the country endpoint
    all_days <- fromJSON(paste("https://api.covid19api.com/total/country", country_slug,sep = "/"))
    
    period <- all_days %>% 
      select(Country, Confirmed, Deaths, Date) %>% 
      mutate(Date = as_date(Date)) %>% 
      filter(Date >= ymd(start)-1  & Date <= end ) %>% 
      mutate(new_case = Confirmed - lag(Confirmed)) %>% 
      mutate(new_death = Deaths - lag(Deaths)) %>% 
      slice(-1L) %>% 
      select(Country,Date,new_case,new_death)
  }

  return(period)
}
```

# Data Exploration
Now that we can interact with a few of the endpoints of the Covid API, let's get some data from them.



